---
title: 'DATA 622: Homework 1'
author: "Andrew Bowen"
date: "2024-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r library, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)
library(recipes)
library(modelr)
```


## Introduction



## Data Wrangling

### NFL Spread Data

I pulled a [dataset from Kaggle here](https://www.kaggle.com/datasets/tobycrabtree/nfl-scores-and-betting-data) containing scores and betting spread data for NFL games dating back to 1967. There's ~15k rows here in this dataset. 

```{r}
nfl_games <- read.csv("data/game_results.csv")

nfl_games$team_away <- as.factor(nfl_games$team_away)
nfl_games$team_home <- as.factor(nfl_games$team_home)

nfl_games$date <- as.Date(nfl_games$schedule_date,  tryFormats = c("%m/%d/%Y"))
```

We can lookup the team IDs from the provided `nfl_teams.csv` file provided by the Kaggle dataset
```{r warning=FALSE, message=FALSE}
teams <- read_csv("data/nfl_teams.csv")

home_teams <- teams %>% rename(home_team_id = team_id) %>% select(team_name, home_team_id) 
away_teams <- teams %>% rename(away_team_id = team_id) %>% select(team_name, away_team_id) 

head(nfl_games)
```

## ELO Ratings
There's also an [ELO dataset for NFL teams dating back to 1920 as well](https://www.kaggle.com/datasets/dtrade84/nfl-elo-ratings). ELO is a [rating system](https://en.wikipedia.org/wiki/Elo_rating_system), originally used to produce the relative ratings of chess players, which can be re-purposed for other competitions. You can read more about [FiveThirtyEight's ELO Rating system here](https://fivethirtyeight.com/methodology/how-our-nfl-predictions-work/). In our case, we'll be looking to see whether ELO ratings are helpful predictors for NFL spread data.
```{r elo-read-in, message=FALSE}
elo <- read_csv("data/nfl_elo.csv")
head(elo)
```

We can filter out any matchups prior to the 1966 season, where our spread data begins
```{r filter-elo}
starting_season <- min(nfl_games$schedule_season)

elo <- elo %>% filter(season > starting_season)

home_teams <- teams %>% rename(home_team_name=team_name) %>% select(home_team_name, team_id) 
away_teams <- teams %>% rename(away_team_name=team_name) %>% select(away_team_name, team_id) 

# Lookup team names based on ID
home_join = join_by(team1 == team_id)
away_join = join_by(team2 == team_id)
elo <- left_join(elo, home_teams, by=home_join) %>% 
  left_join(away_teams, by=away_join)

head(elo)
```


Now we can join together our spread and ELO datasets. We'll need to create a unique identifier for each game (`game_id`), based on the date as well as the teams involved. This should produce a UUID for our dataset
```{r game-ids}
# Create Game ID for joining
nfl_games <- nfl_games %>% mutate(game_id = paste(as.character(date),  team_home, team_away))
elo <- elo %>% mutate(game_id = paste(as.character(date), home_team_name, away_team_name))

# Lookup ELO ratins per game
df <- left_join(nfl_games, elo, by="game_id")

head(df)
```


### Visualizing our NFL Data

```{r homw-elo-vs-score, warning=FALSE}
ggplot(df, aes(x=elo1_pre, y=score_home)) +
  geom_point() +
  labs(x="Home Team ELO Before Game",
       y="Home Team Score",
       title="NFL Home Team Scores vs prior ELO Ratings")
```

```{r away-elo-vs-score, warning=FALSE}
ggplot(df, aes(x=elo2_pre, y=score_away)) + 
  geom_point() + 
  labs(x="Away Team ELO Before Game",
       y="Away Team Score",
       title="NFL Away Team Scores vs prior ELO Ratings")
```



## NBA Boxscores
I also found this other [Kaggle dataset with NBA Boxscores](https://www.kaggle.com/datasets/szymonjwiak/nba-traditional) which would be considered more "big data" (~700k rows).
```{r}

```

